An Intrinsically Typed Solution for the List-Machine Benchmark

### 2020-06-29 papers/SBLP2020/paper.tex
Done some re-reading andfixing some tiny things mainly on comparison section.

\begin{abstract}
Formal models are important tools in the programming languages research
community. However, such models are full of intrincacies and, due to that,
they are subject to subtle errors. Such failures motivated the usage of
tools to ensure the correctness of these formalisms. One way to eliminate
such errors is to encode models in a dependently-typed language in order
to ensure its ``correctness-by-construction''. In this paper, we use this
idea to build a verified interpreter for the list-machine benchmark in the
Agda programming language, comparing the results with formalizations using
the usual syntactic approach. We formalize the 14 tasks of the benchmark
using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
compared to a Coq solution, even without the use of proof automation.
\end{abstract}

@@

The rest of this paper is organized as follows. Section~\ref{sec:agda}
presents a brief introduction to Agda and Section~\ref{sec:list}
reviews the list-machine benchmark and presents its syntax and type system.
We describe the intrinsically-typed representation for the list-machine
in Section~\ref{sec:typing}. The list-machine semantics and its
realization as a definitional interpreter are presented in Section~\ref{sec:semantics}.
The type-checking algorithm, the subtyping relation and the least-common-supertype
algorithm are presented in Section~\ref{sec:typechecker}. Section~\ref{sec:comparison}
compares the presented formalization in Agda with Coq and Twelf encodings. Related work is
discussed in Section~\ref{sec:related}, and Section~\ref{sec:conclusion} concludes.

@@

Each program variable is assigned to a list type, \textcolor{red}{which is used to guarantee the safety of execute
fetch-field that demands non-empty list arguments and branch instructions refine the argument type
according to which branch is taken.} In order to express such refinements, types are subject to a
subtyping relation. The meta-variable $\tau$ denotes an arbitrary type. 
\[
\begin{array}{rcll}
  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
\end{array}
\]
We let notation $\tau \subset \tau'$
denote the subtyping judgement which is defined as follows.
\[
\begin{array}{ccc}
  \inference{}
            {\tau\subset \tau}
            [refl]
  &
  \inference{}
            {nil \subset \tau}
            [nil]
  &
  \inference{\tau \subset \tau'}
            {list\:\tau\subset list\:\tau'}
            [list]\\\\
  \multicolumn{3}{c}{
  \inference{\tau \subset \tau'}
            {listcons\:\tau\subset list\:\tau'}
            [listcons]} \\\\
            \multicolumn{3}{c}{
            \inference{\tau \subset \tau'}
            {listcons\:\tau\subset listcons\:\tau'}
            [mixed]} \\\\
\end{array}
\]
Basically, the subtyping relation specifies that $nil$ (empty list type) is
subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
The other rules specifies that type constructors $list$ and $listcons$ respect
the subtyping relation. The list common supertype $\tau = \tau_1 \sqcap \tau_2$ of
$\tau_1$ and $\tau_2$ is defined as the smallest type such that $\tau_1$ and $\tau_2$
are subtypes of $\tau$.

@@

\[
\begin{array}{cc}
   \inference{}
             {\Pi;\Gamma\vdash_{\text{block}} \textbf{halt}}[halt]
   &

   \inference{\Pi\vdash_{\text{instr}} \Gamma\{\iota_1\}\Gamma'\\ \Pi;\Gamma'\vdash_{\text{block}} \iota_2}
             {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[block-seq]\\ \\

@@

   &
   \inference{}
             {\Pi\vdash_{\text{blocks}} \textbf{end}}[empty] \\ \\

@@

And finally, a \ensuremath{\D{Program}} is a sequence of instruction blocks, each preceded by a label.
We use the \ensuremath{\V{All}} datatype to express this relation.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\D{Program}\;\mathbin{:}\;\D{PCtx}\;\rightarrow\;\D{Set}{}\<[E]%
\\
\>[B]{}\D{Program}\;\V{\Pi}\;\mathrel{=}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{All}\;(\lambda\;\V{\Gamma}\;\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'})\;\V{\Pi}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

We use the datatype \ensuremath{\V{All}} (and \ensuremath{\D{Allv}} for vectors) to define the notion of well-typed
variable environments and well-typed programs. Thus, intuitively, \ensuremath{\D{Env}} is like a
list of well-typed values. And \ensuremath{\D{PEnv}} is like a list of well-typed environments.
Both type environments are used to type block instructions and sequences of
block instructions.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\D{Env}\;\mathbin{:}\;\D{Ctx}\;\rightarrow\;\D{Set}{}\<[E]%
\\
\>[B]{}\D{Env}\;\V{\Gamma}\;\mathrel{=}\;\V{All}\;(\lambda\;(\V{x}\;\red{,}\,\;\V{\tau})\;\rightarrow\;\D{Val}\;\V{\tau})\;\V{\Gamma}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\paragraph{Fetching list information}{The next code shows the evaluation of two syntactical constructors,
both related to the \emph{step-fetch-field-0} rule~\cite{Appel07}. The first retrieves the head element of a list, and
stores it in a new variable. The \ensuremath{\V{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time
environment \ensuremath{\V{env}}, and this variable is added to the result environment. The typed \emph{de Bruijn}
indices guarantee that the projected value has the type demanded, since the environment \ensuremath{\V{env}} is typed
by the context \ensuremath{\V{\Gamma}}. Similarly, the second instruction also retrieves the head element of a list, however
it needs to update the run-time environment on the position of index \ensuremath{\V{v'}}. This process is done by the
function \ensuremath{\F{update\textrm{-}env}}\footnote{The source-code of this function can be found in our online repository.}.}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\V{v})\;\V{b}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathkw{with}\;\V{lookup}\;\V{env}\;\V{v}{}\<[E]%
\\
\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\V{v₁}\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
\\
\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;\V{v}\;\V{v'})\;\V{b}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathkw{with}\;\V{lookup}\;\V{env}\;\V{v}{}\<[E]%
\\
\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v'}\;\V{v₁})\;\V{b}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\paragraph{List creation}{To evaluate the instruction which creates a new list and respect the expected
types, we need some extra lemmas. First because when we create a list from variables \ensuremath{\V{v₀}} and \ensuremath{\V{v₁}}, the
result type of this list is the least common subtype between these two. As before, we use the \ensuremath{\V{lookup}}
function to retrieve the type information using the \emph{de Bruijn} indices of both variables, and we
extend the run-time environment \ensuremath{\V{env}} with the type of the created list. To convince the Agda's
type-checker the new environment is well-typed we use subtyping lemmas, such as \ensuremath{\F{\textrm{<:-}val}} and \ensuremath{\F{list\textrm{-<:-}inv}},
and others lemmas to deal with the least common subtying, such as \ensuremath{\F{lub\textrm{-}subtype\textrm{-}left}}, \ensuremath{\F{lub\textrm{-}subtype\textrm{-}right}},
and \ensuremath{\F{lub\textrm{-}of\textrm{-}subtype}}. These lemmas and their proofs can be found in our repository online. }

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}cons\textrm{-}new}\;\V{v₀}\;\V{v₁}\;\V{s})\;\V{b})\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;((\F{\textrm{<:-}val}\;(\F{list\textrm{-<:-}inv}\;(\F{lub\textrm{-}subtype\textrm{-}left}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}(\F{lub\textrm{-}of\textrm{-}subtype}\;(\F{lub\textrm{-}subtype\textrm{-}left}\;\V{s}))))\;(\V{lookup}\;\V{env}\;\V{v₀}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\V{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\begin{table}[!htb]
\begin{tabular}{rl|rrr}
    & Task                                         & \multicolumn{1}{l}{Twelf} & \multicolumn{1}{l}{Coq} & \multicolumn{1}{l}{Agda} \\ \hline
1.  & Operational Semantics                        & 126                       & 98                      & 109                      \\
2.  & Derive $p \Downarrow$                        & 1                         & 8                       & -                        \\ \hline
3.  & Type System $\vDash_{\textrm{prog}} p : \Pi$ & 167                       & 130                     & 62                       \\
4.  & $\sqcap$ algorithm                           & *                         & *                       & 13                       \\
5.  & Derive $\vDash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
                                                   & 1                         & no                      & -                        \\
6.  & State properties of $\sqcap$                 & 12                        & 13                      & 6                        \\
7.  & Prove properties of $\sqcap$                 & 114                       & 21                      & 124                      \\
8.  & State soundness theorem                      & 29                        & 15                      & *                        \\
9.  & Prove soundness of $\vDash_{\textrm{prog}} p : \Pi$
                                                   & 2060                      & 315                     & *                        \\ \hline
10. & Efficient algorithm                          & 22                        & 145                     & 98                       \\
11. & Derive $\vdash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
                                                   & 1                         & 1                       & 1                        \\
12. & Prove termination of $\vdash_{\textrm{prog}} p : \Pi$
                                                   & 18                        & 0                       & 0                        \\
13. & Scalable type-checker                        & yes                       & yes                     & yes                      \\
14. & Prove soundness of $\vdash_{\textrm{prog}} p : \Pi$
                                                   & 347                       & 141                     & * 
\end{tabular}
\end{table}

@@

\begin{enumerate}
  \item \textbf{Operational semantics.} Instead of using a inductive type for representing the operational semantics, we chose to
        use a definitional interpreter for our intrinsincally-typed representation. Our implementation for the operational
        semantics is composed by 38 lines for the type syntax and 71 for the definitional interpreter.
  \item \textbf{Derive $p \Downarrow$.} Since we used a definitional intepreter for representing the semantics, we can derive
        $p\,\Downarrow$ just by executing the interpreter on $p$.
  \item \textbf{Represent the type system.} Our type system representation consists of the intrinsically-typed encoding of
        the list-machine programs and it is already counted in as part of the operational semantics of our solution.
  \item \textbf{Least common supertype algorithm.} In order to implement this task, we specified the least common super type
        as a relation and implemented the algorithm which shows that this relation is indeed a function. In this way, we
        guarantee its soundness w.r.t. its specification.
  \item \textbf{Derive an example of type checking.} Our approach to build a derivation for a sample list-machine program
        is just execute the type checker over it. Since our type checker returns an intrinsincally-typed representation of
        the input and such typed syntax is equivalent to the type system, it can be used to build type derivations.
  \item \textbf{State properties of the least common supertype.} This is trivial, since it is just a matter to encode the
        desired properties as Agda function types.
  \item \textbf{Prove properties of the least common supertype.} Proofs of all proved properties about the least common supertype
        are simple recursive functions over the least common supertype and subtyping relations definition.
  \item \textbf{State the soundness theorem for the type system.} Following~\citet{Amin17}, we represent the soundness
        theorem statement as the type of our definitional interpreter for the intrinsically-typed syntax for the list-machine
        programs.
  \item \textbf{Prove the soundness theorem for the type system.} Following \citet{Amin17}, the proof of the soundness
        theorem is just the implementation of the definitional interpreter. Soundness is ensured by construction since the interpreter
        produces, as result, an environment of well-typed values resulting from the execution of the input program.
  \item \textbf{Asymptotically efficient algorithm.} Our current implementation of the typechecker takes quadratic time. The reason
        for this inneficiency is the representation of environments as lists / vectors. The use of better data structures (like finite
        maps implemented by efficient search trees) can solve this issue. We leave this fix to future work.
  \item \textbf{Simulate the new algorithm.} This task is just one line of code, since it is only a matter of calling the typechecker
        on the input program.
  \item \textbf{Prove the termination of the type checker.} This task is trivial in our setting, since all defined Agda functions
        must be total. The totality is ensured by Agda's termination / totality checker.
  \item \textbf{Scalable type-checker.} Agda code can be compiled to machine code using its GHC-Haskell backend. Since GHC is
        an industrial strength compiler, the backend can generate an efficient executable for the machine interpreter and typechecker. 
  \item \textbf{Prove soundness of type-checker.} In our approach, the soundness of the typechecker is ensured by construction,
        since it returns the intrinsically-typed representation of the input program which corresponds to its typing derivation.
\end{enumerate}

@@

This paper shows the combination of intrinsically-typed syntax and definitional interpreters to formalize programming languages. Using such approach, we were able to provide a machine checked version of the list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages. The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM, since we were able to encode features such as jumps, mutable state, and sub-typing. When comparing such approach with the conventional syntactic way to prove properties, we can affirm that it requires a small number of lines to achieve the same result, even without the use of proof automation. This happens because the approach uses the power of the host language, and provides some proofs for free due to the intrinsically-typed syntax.

>>>>

\begin{abstract}
Formal models are important tools in the programming languages research
community. However, such models are full of intrincacies and, due to that,
they are subject to subtle errors. Such failures motivated the usage of
tools to ensure the correctness of these formalisms. One way to eliminate
such errors is to encode models in a dependently-typed language in order
to ensure its ``correctness-by-construction''. In this paper, we use this
idea to build a verified interpreter for the list-machine benchmark in the
Agda programming language, comparing the results with formalizations developed
by Appel and Leroy. We formalize the 14 tasks of the benchmark
using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
compared to a Coq solution, even without the use of proof automation.
\end{abstract}

@@

The rest of this paper is organized as follows. Section~\ref{sec:agda}
presents a brief introduction to Agda and Section~\ref{sec:list}
reviews the list-machine benchmark and presents its syntax and type system.
We describe the intrinsically-typed representation for the list-machine
in Section~\ref{sec:typing}. Section~\ref{sec:semantics} briefly discuss
the list-machine semantics and then proceeds with its
realization as a definitional interpreter.
The type-checking algorithm, the subtyping relation and the least-common-supertype
algorithm are presented in Section~\ref{sec:typechecker}. Section~\ref{sec:comparison}
compares the presented formalization in Agda with Coq and Twelf encodings. Related work is
discussed in Section~\ref{sec:related}, and Section~\ref{sec:conclusion} concludes.

@@

Each program variable is assigned to a list type, which is used to guarantee the safety of execute
fetch-field that demands non-empty list arguments. In order to express such refinements, types are subject to a
subtyping relation. The meta-variable $\tau$ denotes an arbitrary type. 
\[
\begin{array}{rcll}
  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
\end{array}
\]
We let notation $\tau \subset \tau'$
denote the subtyping judgement which is defined as follows.
\[
\begin{array}{ccc}
  \inference{}
            {\tau\subset \tau}
            [refl]
  &
  \inference{}
            {nil \subset \tau}
            [nil]
  &
  \inference{\tau \subset \tau'}
            {list\:\tau\subset list\:\tau'}
            [list]\\\\
  \multicolumn{3}{c}{
  \inference{\tau \subset \tau'}
            {listcons\:\tau\subset list\:\tau'}
            [listcons]} \\\\
            \multicolumn{3}{c}{
            \inference{\tau \subset \tau'}
            {listcons\:\tau\subset listcons\:\tau'}
            [mixed]} \\\\
\end{array}
\]
Basically, the subtyping relation specifies that $nil$ (empty list type) is
subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
The other rules specifies that type constructors $list$ and $listcons$ respect
the subtyping relation. The list common supertype $\tau = \tau_1 \sqcap \tau_2$ of
$\tau_1$ and $\tau_2$ is defined as the smallest type such that $\tau_1$ and $\tau_2$
are subtypes of $\tau$.

@@

\[
\begin{array}{cc}
   \inference{}
             {\Pi;\Gamma\vdash_{\text{block}} \textbf{halt}}[halt]
   &
   \inference{}
             {\Pi\vdash_{\text{blocks}} \textbf{end}}[empty]\\ \\

@@

   &

   \inference{\Pi\vdash_{\text{instr}} \Gamma\{\iota_1\}\Gamma'\\ \Pi;\Gamma'\vdash_{\text{block}} \iota_2}
             {\Pi;\Gamma\vdash_{\text{block}} \iota_1;\iota_2}[block-seq]
 \\ \\

@@

And finally, a \ensuremath{\D{Program}} is a sequence of instruction blocks, each preceded by a label.
We use the \ensuremath{\D{All}} datatype to express this relation.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\D{Program}\;\mathbin{:}\;\D{PCtx}\;\rightarrow\;\D{Set}{}\<[E]%
\\
\>[B]{}\D{Program}\;\V{\Pi}\;\mathrel{=}\;\D{\forall}\;\{\mskip1.5mu \V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\D{All}\;(\lambda\;\V{\Gamma}\;\rightarrow\;\D{Block}\;\V{\Pi}\;\V{\Gamma}\;\V{\Gamma'})\;\V{\Pi}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

We use the datatype \ensuremath{\D{All}} (and \ensuremath{\D{Allv}} for vectors) to define the notion of well-typed
variable environments and well-typed programs. Thus, intuitively, \ensuremath{\D{Env}} is like a
list of well-typed values. And \ensuremath{\D{PEnv}} is like a list of well-typed environments.
Both type environments are used to type block instructions and sequences of
block instructions.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\D{Env}\;\mathbin{:}\;\D{Ctx}\;\rightarrow\;\D{Set}{}\<[E]%
\\
\>[B]{}\D{Env}\;\V{\Gamma}\;\mathrel{=}\;\D{All}\;(\lambda\;(\V{x}\;\red{,}\,\;\V{\tau})\;\rightarrow\;\D{Val}\;\V{\tau})\;\V{\Gamma}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\paragraph{Fetching list information}{The next code shows the evaluation of two syntactical constructors,
both related to the \emph{step-fetch-field-0} rule~\cite{Appel07}. The first retrieves the head element of a list, and
stores it in a new variable. The \ensuremath{\F{lookup}} function projects the value of variable \ensuremath{\V{v}} from the run-time
environment \ensuremath{\V{env}}, and this variable is added to the result environment. The typed \emph{de Bruijn}
indices guarantee that the projected value has the type demanded, since the environment \ensuremath{\V{env}} is typed
by the context \ensuremath{\V{\Gamma}}. Similarly, the second instruction also retrieves the head element of a list, however
it needs to update the run-time environment on the position of index \ensuremath{\V{v'}}. This process is done by the
function \ensuremath{\F{update\textrm{-}env}}\footnote{The source-code of this function can be found in our online repository.}.}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}new}\;\V{v})\;\V{b}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
\\
\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\V{v₁}\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
\\
\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}fetch\textrm{-}0\textrm{-}upd}\;\V{v}\;\V{v'})\;\V{b}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
\\
\>[B]{}\V{...|}\;\V{v₁}\;\Con{::}\;\V{v₂}\;\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{v'}\;\V{v₁})\;\V{b}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\paragraph{List creation}{To evaluate the instruction which creates a new list and respect the expected
types, we need some extra lemmas. First because when we create a list from variables \ensuremath{\V{v₀}} and \ensuremath{\V{v₁}}, the
result type of this list is the least common subtype between these two. As before, we use the \ensuremath{\F{lookup}}
function to retrieve the type information using the \emph{de Bruijn} indices of both variables, and we
extend the run-time environment \ensuremath{\V{env}} with the type of the created list. To convince the Agda's
type-checker the new environment is well-typed we use subtyping lemmas, such as \ensuremath{\F{\textrm{<:-}val}} and \ensuremath{\F{list\textrm{-<:-}inv}},
and others lemmas to deal with the least common subtying, such as \ensuremath{\F{lub\textrm{-}subtype\textrm{-}left}}, \ensuremath{\F{lub\textrm{-}subtype\textrm{-}right}},
and \ensuremath{\F{lub\textrm{-}of\textrm{-}subtype}}. These lemmas and their proofs can be found in our repository online. }

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}cons\textrm{-}new}\;\V{v₀}\;\V{v₁}\;\V{s})\;\V{b})\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{p}\;((\F{\textrm{<:-}val}\;(\F{list\textrm{-<:-}inv}\;(\F{lub\textrm{-}subtype\textrm{-}left}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}(\F{lub\textrm{-}of\textrm{-}subtype}\;(\F{lub\textrm{-}subtype\textrm{-}left}\;\V{s}))))\;(\F{lookup}\;\V{env}\;\V{v₀}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{::}\;\F{\textrm{<:-}val}\;(\F{lub\textrm{-}subtype\textrm{-}right}\;\V{s})\;(\F{lookup}\;\V{env}\;\V{v₁}))\;\Con{::}\;\V{env})\;\V{b}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\begin{table}[!htb]
\begin{tabular}{rl|rrr}
    & Task                                         & \multicolumn{1}{l}{Twelf} & \multicolumn{1}{l}{Coq} & \multicolumn{1}{l}{Agda} \\ \hline
1.  & Operational Semantics                        & 126                       & 98                      & 109                      \\
2.  & Derive $p \Downarrow$                        & 1                         & 8                       & 1                        \\ \hline
3.  & Type System $\vDash_{\textrm{prog}} p : \Pi$ & 167                       & 130                     & 62                       \\
4.  & $\sqcap$ algorithm                           & *                         & *                       & 13                       \\
5.  & Derive $\vDash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
                                                   & 1                         & no                      & 1                        \\
6.  & State properties of $\sqcap$                 & 12                        & 13                      & 6                        \\
7.  & Prove properties of $\sqcap$                 & 114                       & 21                      & 124                      \\
8.  & State soundness theorem                      & 29                        & 15                      & *                        \\
9.  & Prove soundness of $\vDash_{\textrm{prog}} p : \Pi$
                                                   & 2060                      & 315                     & *                        \\ \hline
10. & Efficient algorithm                          & 22                        & 145                     & 98                       \\
11. & Derive $\vdash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
                                                   & 1                         & 1                       & 1                        \\
12. & Prove termination of $\vdash_{\textrm{prog}} p : \Pi$
                                                   & 18                        & 0                       & 0                        \\
13. & Scalable type-checker                        & yes                       & yes                     & yes                      \\
14. & Prove soundness of $\vdash_{\textrm{prog}} p : \Pi$
                                                   & 347                       & 141                     & * 
\end{tabular}
\end{table}

@@

\begin{enumerate}
  \item \textbf{Operational semantics.} Instead of using a inductive type for representing the operational semantics, we chose to
        use a definitional interpreter for our intrinsincally-typed representation. Our implementation for the operational
        semantics is composed by 38 lines for the type syntax and 71 for the definitional interpreter.
  \item \textbf{Derive $p \Downarrow$.} Since we used a definitional intepreter for representing the semantics, we can derive
        $p\,\Downarrow$ just by executing the interpreter on $p$.
  \item \textbf{Represent the type system.} Our type system representation consists of the intrinsically-typed encoding of
        the list-machine programs and it is already counted in as part of the operational semantics of our solution.
  \item \textbf{Least common supertype algorithm.} In order to implement this task, we specified the least common super type
        as a relation and implemented the algorithm which shows that this relation is indeed a function. In this way, we
        guarantee its soundness w.r.t. its specification.
  \item \textbf{Derive an example of type checking.} Our approach to build a derivation for a sample list-machine program
        is just execute the type checker over it. Since our type checker returns an intrinsincally-typed representation of
        the input and such typed syntax is equivalent to the type system, it can be used to build type derivations.
  \item \textbf{State properties of the least common supertype.} This is trivial, since it is just a matter to encode the
        desired properties as Agda function types.
  \item \textbf{Prove properties of the least common supertype.} Proofs of all proved properties about the least common supertype
        are simple recursive functions over the least common supertype and subtyping relations definitions and were omitted
        from this text for space reasons.
  \item \textbf{State the soundness theorem for the type system.} Following~\citet{Amin17}, we represent the soundness
        theorem statement as the type of our definitional interpreter for the intrinsically-typed syntax for the list-machine
        programs. Therefore, this was done as part of task 9.
  \item \textbf{Prove the soundness theorem for the type system.} Following \citet{Amin17}, the proof of the soundness
        theorem is just the implementation of the definitional interpreter. Soundness is ensured by construction since the interpreter
        produces, as result, an environment of well-typed values resulting from the execution of the input program.
  \item \textbf{Asymptotically efficient algorithm.} Our current implementation of the typechecker takes quadratic time. The reason
        for this inneficiency is the representation of environments as lists / vectors. The use of better data structures (like finite
        maps implemented by efficient search trees) can solve this issue. We leave this fix to future work.
  \item \textbf{Simulate the new algorithm.} This task is just one line of code, since it is only a matter of calling the typechecker
        on the input program.
  \item \textbf{Prove the termination of the type checker.} This task is trivial in our setting, since all defined Agda functions
        must be total. The totality is ensured by Agda's termination / totality checker.
  \item \textbf{Scalable type-checker.} Agda code can be compiled to machine code using its GHC-Haskell backend. Since GHC is
        an industrial strength compiler, the backend can generate an efficient executable for the machine interpreter and typechecker. 
  \item \textbf{Prove soundness of type-checker.} In our approach, the soundness of the typechecker is ensured by construction,
        since it returns the intrinsically-typed representation of the input program which corresponds to its typing derivation.
\end{enumerate}

@@

This paper shows how the combination of intrinsically-typed syntax and definitional interpreters can be used to simplify
programming languages formalization tasks. Using such approach, we were able to provide a machine checked version of the
list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages.
The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM,
since we were able to encode features such as jumps, mutable state, and sub-typing. When comparing such our work with the
conventional non-dependently typed formalization strategies (like to ones used by Appel et. al. in his Coq implementation),
we can affirm that it requires a small number of lines to achieve the same result, even without the use of proof automation.
This happens because the approach uses the power of the host language, and provides some proofs for free due to the
intrinsically-typed syntax.
